Supplemental Online Materials:\\\textit{Double Standards in Judging Collective Action}

### 2024-04-16 Documents/supplemental-materials.tex
resubmit

We estimated a graded response model (Bürkner, 2021; Samejima, 1997), an
item response theory model for ordinal response variables, for
participants' ratings of how often an action would be an acceptable form
of collective action. For each participant \(j\), the model estimated
their unique propensity (\(\theta_j\)) to consider collective actions
acceptable in more situations. For each item \(i\), the model estimated
four acceptability thresholds (\(\beta_{ik}\)), separating the five
response options, and one discrimination parameter (\(\alpha_i\))
indicating how well the item differentiates between participants with
different propensities to consider collective actions acceptable. We
estimated the model in \emph{CmdStanR} using similar prior distribution
as in Experiment. We used an induced Dirichlet prior (Betancourt, 2019)
for the item-specific difficulty thresholds.

Table S1 shows each actions' item information and discrimination as well
as proportions for the observed responses. Table S2 shows participants'
averaged ratings of how acceptable, disruptive, violent, extreme, and
negative they considered each action to be. Table S3 shows that all five
ratings were strongy correlated across actions and participants. We
considered both estimates from the graded response model and their
relevance to the social contexts to select actions for Experiments 1 and
2.

@@

@@

@@

\begin{table}[t!]
\caption{Correlations between variables in Experiment 1}
\centering
\begin{tabular}{rlrrrr}
\toprule
\# & Variable              &    1 &    2 &    3 &    4 \\ \midrule
 1 & Proportion            &      & -.31 & -.16 & -.23 \\
 2 & Political Orientation & -.31 &      &  .41 &  .41 \\
 3 & Social Dominance      & -.16 &  .41 &      &  .29 \\
 4 & System Justification  & -.23 &  .41 &  .29 &      \\ \bottomrule
\addlinespace
\end{tabular}
\caption*{\textit{Note.} Proportion of actions considered acceptable (0--1); Political Orientation (1 = \textit{left}, 7 = \textit{right}); Social Dominance Orientation (\textit{factor score}); Systen-Justifying Beliefs (\textit{factor score})}
\end{table}

@@

\begin{table}[t!]
\caption{Correlations between variables in Experiment 2}
\centering
\begin{tabular}{rlrrrrr}
\toprule
\# & Variable              &    1 &    2 &    3 &    4 &    5 \\ \midrule
 1 & Proportion            &      & -.21 &  .29 & -.23 & -.07 \\
 2 & System Justification  & -.21 &      & -.45 &  .47 &  .04 \\
 3 & Policy Support        &  .29 & -.45 &      & -.49 &  .00 \\
 4 & Political Orientation & -.23 &  .47 & -.49 &      &  .03 \\
 5 & Identification        & -.07 &  .04 &  .00 &  .03 &      \\ \bottomrule
\addlinespace
\end{tabular}
\caption*{\textit{Note.} Proportion of actions considered acceptable (0--1); Systen-Justifying Beliefs (\textit{factor score}); Policy Support (1 = \textit{strongly oppose}, 5 = \textit{strongly support}); Political Orientation (1 = \textit{liberal}, 7 = \textit{conservative}); Racial Identification (1 = \textit{not at all}, 7 = \textit{very much})}
\end{table}

@@

\begin{table}[t!]
\caption{Correlations between variables in Experiment 3}
\centering
\begin{tabular}{rlrrrrr}
\toprule
\# & Variable              &    1 &    2 &    3 &    4 &    5 \\ \midrule
 1 & Proportion            &      & -.27 & -.22 & -.32 & -.11 \\
 2 & System Justification  & -.27 &      &  .61 &  .78 &  .16 \\
 3 & Policy Support        & -.22 &  .61 &      &  .74 &  .16 \\
 4 & Political Orientation & -.32 &  .78 &  .74 &      &  .20 \\
 5 & Identification        & -.11 &  .16 &  .16 &  .20 &      \\ \bottomrule
\addlinespace
\end{tabular}
\caption*{\textit{Note.} Proportion of actions considered acceptable (0--1); Systen-Justifying Beliefs (\textit{factor score}); Policy Support (1 = \textit{strongly oppose}, 5 = \textit{strongly support}); Political Orientation (1 = \textit{liberal}, 7 = \textit{conservative}); Gender Identification (1 = \textit{not at all}, 7 = \textit{very much})}
\end{table}

>>>>

We estimated a graded response model (Bürkner, 2021; Samejima, 1997), an
item response theory model for ordinal response variables, for
participants' ratings of how often an action would be an acceptable form
of collective action. For each participant \(j\), the model estimated
their unique propensity (\(\theta_j\)) to consider collective actions
acceptable in more situations. For each item \(i\), the model estimated
four acceptability thresholds (\(\beta_{ik}\)), separating the five
response options, and one discrimination parameter (\(\alpha_i\))
indicating how well the item differentiates between participants with
different propensities to consider collective actions acceptable. We
estimated the model in \emph{CmdStanR} using similar prior distribution
as in Experiment 1. We used an induced Dirichlet prior (Betancourt,
2019) for the item-specific difficulty thresholds.

Table S1 shows each protest action's information and discrimination
parameters as well as the proportions of participants choosing each
answer option. Table S2 shows participants' averaged ratings of how
acceptable, disruptive, violent, extreme, and negative they considered
each action to be. Table S3 shows that all five ratings were strongly
correlated across actions and participants.

Figure \ref{fig:s1} shows the scale information curves for scales
composed of, respectively, the 1--72 most informative and discriminating
protest actions. We selected the 25 most informative actions to
construct the final scale to be used in Experiments 1--3. While adding
more actions to the scale would have improved the information provided
by the instrument, the added benefit of each additional action included
in the scale would have been small (see Figure \ref{fig:s1}). We limited
the number of protest actions to 25 so as to not overwhelm participants
and to prevent satisficing. Tables A1, B1, and C1 list the protest
actions used in Experiments 1--3.

\begin{figure*}[!t]
\centering
\caption{Scale information curves for scales composed of the 1--72 most informative and discriminating protest actions}
\includegraphics[scale=1]{../Scale Development/figures/figure-s1}
\caption*{\textit{Note.} The information curve for the final scale of 25 actions is highlighted in red. As in Figure 1, $\theta_{(z)}$ follows a standard normal distribution and is shown as both $z$-scores (bottom) and percentiles (top).}
\label{fig:s1}
\end{figure*}

@@

\hypertarget{reanalyses-without-exclusions}{%
\section{Reanalyses Without
Exclusions}\label{reanalyses-without-exclusions}}

We replicated the analyses for Experiments 1-3 using the complete data
without excluding participants based on the (preregistered) exclusion
criteria.

\hypertarget{experiment-1-1}{%
\subsection{Experiment 1}\label{experiment-1-1}}

\begin{figure*}[!t]
\centering
\caption{Results from the preregistered (\textbf{A}) and non-preregistered (\textbf{B}) analyses for Experiment 1 without exclusions}
\includegraphics[scale=1]{../Experiment 1/figures/figure-s2}
\caption*{\textit{Note.} HS = Higher Status, LS = Lower Status. $\theta_{(z)}$ is the \textit{z}-standardized tendency to consider more controversial actions acceptable means of protest. (\textbf{A}) $p$ and $n$ are, respectively, the predicted proportion and number of actions a participant would consider acceptable means of protest in each condition. Bars enclose the 95\% most plausible estimates. (\textbf{B}) Ribbons enclose, from the darkest to the lightest shade, the 50\%, 80\%, 95\%, and 99\% most plausible estimates.}
\label{fig:s2}
\end{figure*}

Figure \ref{fig:s2} (A) shows estimates for each combination of the
protesters' and the participants' group membership. Overall, we found
that participants' responses depended on both the protesters' and the
participants' group status---but not in the directions predicted by our
hypotheses. Contradicting our first hypothesis, participants did not
consider protest actions performed by their ingroup to be more
acceptable, on average, than the same actions performed by the relevant
outgroup (Cohen's \(d = -0.01, [-0.20, 0.16]\)). Instead, we found that
participants from higher-status backgrounds considered protest actions
performed by both lower-status (\(d = 0.40, [0.15, 0.66\){]}) and
higher-status (\(d = 0.37, [0.10, 0.63]\)) protesters to be, on average,
more acceptable than participants from lower-status backgrounds.
Contradicting our second hypothesis, participants considered protest
actions performed by higher-status protesters to be less acceptable, on
average, than the same actions performed by lower-status protesters
(\(d = -0.20, [-0.37, -0.02]\)).

Figure \ref{fig:s2} shows (B) the \(z\)-standardized propensity to
consider more controversial protest actions acceptable as a function of
the three ideological orientation variables. We found that participants
who reported a more right-wing political orientation
(\(\beta_{xy} = -0.24, [-0.34, -0.15]\)) and who expressed more
agreement with system-justifying beliefs
(\(\beta_{xy} = -0.21, [-0.31, -0.12]\)) tended to find fewer collective
actions to be acceptable. In contrast, we found that, after controlling
for the other two variables, social dominance orientation was not
associated with participants' judgments about how acceptable various
collective actions are (\(\beta_{xy} = -0.06, [-0.16, 0.03]\)). Overall,
these findings suggest that people who are right-wing and endorse
system-justifying beliefs tend to find various collective actions to be
less acceptable than people who are left-wing and do not endorse
system-justifying beliefs.

\hypertarget{experiment-2}{%
\subsection{Experiment 2}\label{experiment-2}}

\hypertarget{preregistered-analyses}{%
\subsubsection{Preregistered Analyses}\label{preregistered-analyses}}

\begin{figure*}[!t]
\centering
\caption{Results from the preregistered analyses for Experiment 2}
\includegraphics[scale=1]{../Experiment 2/figures/figure-s3}
\caption*{\textit{Note.} Against = Protesters oppose defunding the police. For = Protesters support defunding the police. $\theta_{(z)}$ is the \textit{z}-standardized tendency to consider more controversial actions acceptable means of protest. Ribbons enclose, from the darkest to the lightest shade, the 50\%, 80\%, 95\%, and 99\% most plausible estimates.}
\label{fig:s3}
\end{figure*}

\begin{figure*}[!t]
\centering
\caption{Predictions from the preregistered (system justification) and non-preregistered (political orientation) analyses for Experiment 2 without exclusions}
\includegraphics[scale=1]{../Experiment 2/figures/figure-s4}
\caption*{\textit{Note.} Against = Protesters oppose defunding the police. For = Protesters support defunding the police. $\theta_{(z)}$ is the \textit{z}-standardized tendency to consider more controversial actions acceptable means of protest.}
\label{fig:s4}
\end{figure*}

Figure \ref{fig:s3} shows results from three preregistered models we
estimated to test our hypotheses.

Model 1 estimated varying intercepts for the eight conditions to test
whether, in line with Hypotheses 1a, 1b, or 2a, participants' responses
depended on their own group membership, the protesters' group
membership, or the protesters' cause. Contradicting \emph{Hypothesis
1a}, participants did not consider protest actions performed by their
ingroup to be more acceptable, on average, than the same actions
performed by the relevant outgroup (\(d = 0.01, [-0.03, 0.06]\)).
Contradicting \emph{Hypothesis 1b}, participants did not consider
protest actions for a cause that was nominally aligned with their
ingroup's interests to be more acceptable, on average, than protest
actions for a cause not aligned with their ingroup's interests
(\(d = -0.01, [-0.06, 0.03]\)). Contradicting \emph{Hypothesis 2a},
participants did not consider protest actions for a system-defending
cause to be more acceptable, on average, than protest actions for a
system-challenging cause (\(d = -0.14, [-0.19, -0.08]\)). Instead, we
found that, on average, White participants considered all protest
actions to be more acceptable than Black participants
(\(d = 0.09, [0.04, 0.14]\)) and both Black (\(d = 0.12, [0.05, 0.19]\))
and White (\(d = 0.15, [0.08, 0.22]\)) participants considered the same
actions to be more acceptable when protests were for, rather than
against, defunding the police. As in Experiment 1, we thus found that
participants' responses depended on both the participants' and the
protesters' group memberships---but not in the directions predicted by
our hypotheses.

Model 2 extended Model 1 by estimating participants' responses as a
function of their \emph{z}-standardized endorsement of system-justifying
beliefs. As preregistered, we modeled this relationship with two fixed
effects, one estimating the effect of system-justifying beliefs on
judgments about system-defending protest actions and one estimating
their effect on judgments about system-challenging protest actions, and
one varying effect estimating its variance across conditions. Supporting
\emph{Hypothesis 2b}, participants who \emph{rejected} system-justifying
beliefs were \emph{more} likely to consider system-challenging protest
actions (for defunding the police) to be acceptable means of protest
(\(\beta_{xy} = 0.40, [0.20, 0.57]\)). We did not, however, find
evidence for the ideological symmetry implied by \emph{Hypothesis 2b}
since participants who \emph{endorsed} system-justifying beliefs were
\emph{not} more likely to consider system-defending protest actions
(against defunding the police) to be acceptable means of protest
(\(\beta_{xy} = -0.17, [-0.37, -0.01]\)).

Figure \ref{fig:s4} shows the estimated pattern of condition-wise
differences underlying those fixed effects. Participants who endorsed
system-justifying beliefs tended to consider system-challenging and
system-defending protest actions to be equally unacceptable. In
contrast, participants who rejected system-justifying beliefs evinced
ideology-based double standards: In line with \emph{Hypothesis 2b}, they
considered the same protest actions to be more acceptable when the
protesters challenged the system or, to a lesser extent, when the
protesters were from the disadvantaged group.

Model 3 extended Model 1 by estimating participants' responses as a
function of their self-reported support for the cause of the protest. We
recoded participants' responses to create a predictor variable that
encoded support for defunding the police when protesters supported
defunding the police and opposition to defunding the police when
protesters opposed defunding the police. As preregistered, we modeled
this relationship as a monotonic effect (Bürkner \& Charpentier, 2020)
that estimated the average change in the outcome variable across
predictor categories as well as how much of this change occurred between
each of the four pairs of adjacent predictor categories.\footnote{Our
  model assigned a Dirichlet prior, \(\alpha = {1, 1, 1, 1}\), to the
  proportions of the overall change that was expected to occur between
  each of the four pairs of predictor categories.} Contradicting
\emph{Hypothesis 3}, participants who supported the protesters' cause
did not, on average, consider the same protest actions to be more
acceptable than participants who opposed the protesters' cause
(\(\beta_{y} = 0.14, [-0.10, 0.39]\)). Ergo, our results did not support
the alternative hypothesis that ideology-based double standards can be
reduced to support for, or opposition to, the cause of a protest.

\hypertarget{non-preregistered-analyses}{%
\subsubsection{Non-Preregistered
Analyses}\label{non-preregistered-analyses}}

We explored whether group identification moderated how the participants'
and the protesters' group memberships affected participants' responses.
To that end, we extended Model 1 by estimating participants' responses
as a function of their \emph{z}-standardized identification with their
racial ingroup. As in Model 2, we modeled this relationship with a fixed
and a varying effect. We found, however, that even participants who
strongly identified with their racial ingroup (+1\emph{SD}) did not
consider protest actions performed by their ingroup to be more
acceptable than the same actions performed by the relevant outgroup
(\(d = 0.06, [-0.01, 0.13]\)) and did not consider protest actions for a
cause that was aligned with their ingroup's interests to be more
acceptable than protest actions for a cause not aligned with their
ingroup's interests (\(d = -0.00, [-0.07, 0.07]\)). Our
non-preregistered analyses thus suggested that group identification did
not moderate group differences in judgments about collective action by
ingroup and outgroup members.

We also explored whether we would find ideology-based double standards
when operationalizing ideology as political orientation instead of as
system justification. To that end, we reran Model 2 with political
orientation as the \emph{z}-standardized predictor variable. Our results
mirrored the preregistered analyses: More liberal participants were more
likely to consider protest actions for defunding the police to be
acceptable (\(\beta_{xy} = 0.40, [0.20, 0.58]\)) but more conservative
participants were not more likely to consider protest actions against
defunding the police to be acceptable
(\(\beta_{xy} = -0.16, [-0.35, 0.01]\)). As Figure \ref{fig:s4} shows,
conservative participants tended to consider all protest actions to be
equally unacceptable while liberal participants considered the same
protest actions to be more acceptable when protesters rallied around a
progressive cause. Our non-preregistered analyses thus replicated the
ideology-based double standards from the preregistered analyses with a
different operationalization of ideology.

\hypertarget{experiment-3-1}{%
\subsection{Experiment 3}\label{experiment-3-1}}

\begin{figure*}[!t]
\centering
\caption{Results from the preregistered analyses for Experiment 3 without exclusions}
\includegraphics[scale=1]{../Experiment 3/figures/figure-s5}
\caption*{\textit{Note.} Results show the estimated \textit{z}-standardized tendency ($\theta_{(z)}$) to consider more controversial actions acceptable means to protest for or against restricting abortion as a function of the participants' ideology and gender.}
\label{fig:s5}
\end{figure*}

\hypertarget{preregistered-analyses-1}{%
\subsubsection{Preregistered Analyses}\label{preregistered-analyses-1}}

Figure \ref{fig:s5} shows the results of our preregistered analyses
without exclusions.

As in Experiments 1 and 2, we found evidence for ideology-based double
standards: Conservative participants considered the same protest actions
to be more acceptable, on average, when protesters \emph{supported}
restricting abortion (\(d = 0.21, [0.03, 0.39]\)) while liberal
participants considered the same protest actions to be more acceptable
when protesters \emph{opposed} restricting abortion
(\(d = 0.87, [0.70, 1.06]\)). This double standard was more pronounced
among liberal participants (\(d = 0.66, [0.42, 0.93]\)). In this way,
Experiment 3 provided evidence for both ideological symmetry and
asymmetry in judging collective action for and against restricting
abortion.

We found some evidence for identity-based double standards: Female
participants considered protest actions \emph{against} restricting
abortion to be more acceptable, on average, than male participants or
protest actions \emph{for} restricting abortion
(\(d = 0.14, [-0.01, 0.29]\)). This pattern was, however, overshadowed
by stronger ideology-based double standards that were consistent across
subsamples: Conservative women (\(d = 0.24, [-0.01, 0.48]\)) and, to a
lesser extent, conservative men (\(d = 0.18, [-0.08, 0.44]\)) considered
the same protest actions to be more acceptable when protesters
\emph{supported} restricting abortion and both liberal women
(\(d = 1.00, [0.73, 1.27]\)) and liberal men
(\(d = 0.75, [0.51, 1.01]\)) considered the same protest actions to be
more acceptable when protesters \emph{opposed} restricting abortion. In
this way, Experiment 3 provided more consistent evidence for
ideology-based than for identity-based double standards in judging
collective action.

\hypertarget{non-preregistered-analyses-1}{%
\subsubsection{Non-Preregistered
Analyses}\label{non-preregistered-analyses-1}}

We explored whether group identification moderated participants'
judgments. We found, however, that gender identification neither
affected how women judged protesters opposing
(\(\beta_{xy} = -0.06, [-0.15, 0.04]\)) or supporting
(\(\beta_{xy} = -0.07, [-0.16, 0.02]\)) abortion restrictions nor how
men judged protesters opposing (\(\beta_{xy} = -0.06, [-0.15, 0.03]\))
or supporting (\(\beta_{xy} = -0.06, [-0.14, 0.03]\)) abortion
restrictions. Our non-preregistered analyses thus suggested that gender
identification did not moderate identity-based double standards in
judging collective action concerning reproductive rights.

@@

\leavevmode\vadjust pre{\hypertarget{ref-burkner_modelling_2020}{}}%
Bürkner, P.-C., \& Charpentier, E. (2020). Modelling monotonic effects
of ordinal predictors in {Bayesian} regression models. \emph{British
Journal of Mathematical and Statistical Psychology}, \emph{73}(3),
420--451. \url{https://doi.org/10.1111/bmsp.12195}

@@

\begin{table}[t!]
\caption{Correlations between variables in Experiment 1}
\centering
\begin{tabular}{rlrrrr}
\toprule
\# & Variable              &    1 &    2 &    3 &    4 \\ \midrule
 1 & Proportion            &      & -.31 & -.16 & -.23 \\
 2 & Political Orientation & -.31 &      &  .41 &  .41 \\
 3 & Social Dominance      & -.16 &  .41 &      &  .29 \\
 4 & System Justification  & -.23 &  .41 &  .29 &      \\ \bottomrule
\addlinespace
\end{tabular}
\caption*{\textit{Note.} Proportion of actions considered acceptable (0--1); Political Orientation (1 = \textit{left}, 7 = \textit{right}); Social Dominance Orientation (\textit{factor score}); System-Justifying Beliefs (\textit{factor score})}
\end{table}

@@

\begin{table}[t!]
\caption{Correlations between variables in Experiment 2}
\centering
\begin{tabular}{rlrrrrr}
\toprule
\# & Variable              &    1 &    2 &    3 &    4 &    5 \\ \midrule
 1 & Proportion            &      & -.21 &  .29 & -.23 & -.07 \\
 2 & System Justification  & -.21 &      & -.45 &  .47 &  .04 \\
 3 & Policy Support        &  .29 & -.45 &      & -.49 &  .00 \\
 4 & Political Orientation & -.23 &  .47 & -.49 &      &  .03 \\
 5 & Identification        & -.07 &  .04 &  .00 &  .03 &      \\ \bottomrule
\addlinespace
\end{tabular}
\caption*{\textit{Note.} Proportion of actions considered acceptable (0--1); System-Justifying Beliefs (\textit{factor score}); Policy Support (1 = \textit{strongly oppose}, 5 = \textit{strongly support}); Political Orientation (1 = \textit{liberal}, 7 = \textit{conservative}); Racial Identification (1 = \textit{not at all}, 7 = \textit{very much})}
\end{table}

@@

\begin{table}[t!]
\caption{Correlations between variables in Experiment 3}
\centering
\begin{tabular}{rlrrrrr}
\toprule
\# & Variable              &    1 &    2 &    3 &    4 &    5 \\ \midrule
 1 & Proportion            &      & -.27 & -.22 & -.32 & -.11 \\
 2 & System Justification  & -.27 &      &  .61 &  .78 &  .16 \\
 3 & Policy Support        & -.22 &  .61 &      &  .74 &  .16 \\
 4 & Political Orientation & -.32 &  .78 &  .74 &      &  .20 \\
 5 & Identification        & -.11 &  .16 &  .16 &  .20 &      \\ \bottomrule
\addlinespace
\end{tabular}
\caption*{\textit{Note.} Proportion of actions considered acceptable (0--1); System-Justifying Beliefs (\textit{factor score}); Policy Support (1 = \textit{strongly oppose}, 5 = \textit{strongly support}); Political Orientation (1 = \textit{liberal}, 7 = \textit{conservative}); Gender Identification (1 = \textit{not at all}, 7 = \textit{very much})}
\end{table}
